---
title: '#TidyTuesday - Beach Volleyball'
author: "Evan Canfield"
date: "5/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Link**: [https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-19/readme.md](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-19/readme.md)

# Load
## Libraries
```{r load_library}
if (!require(pacman)) {install.packages('pacman')} 
p_load(
  ggridges,
  lubridate,
  skimr,
  stringr,
  tidyverse,
  viridis
)
```

## Data
```{r load_data}
vb_matches <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-19/vb_matches.csv', guess_max = 76000)
```

# EDA
```{r}
vb_matches %>% head(10)
```

```{r}
vb_matches %>% skim()
```
# Transformation
I need to calculate the average player height and standard deviation to create a baseline.

```{r}
hgt_avg <- vb_matches %>% 
  select(w_p1_hgt, w_p2_hgt, l_p1_hgt, l_p2_hgt) %>% 
  pivot_longer(cols = everything(),
              names_to = "player", 
              values_to = "height") %>%
  drop_na() %>% 
  summarise(hgt_avg = mean(height)) %>% 
  as.numeric()

hgt_avg

hgt_sd <- vb_matches %>% 
  select(w_p1_hgt, w_p2_hgt, l_p1_hgt, l_p2_hgt) %>% 
  pivot_longer(cols = everything(),
              names_to = "player", 
              values_to = "height") %>%
  drop_na() %>% 
  summarise(hgt_sd = sd(height)) %>% 
  as.numeric()

hgt_sd
```

Now, I'll calculate the average height of each team in each match, and select important grouping variables. We will also calculate if the taller team won, and if the team was 1 standard deviation taller and won.
```{r}
df <- vb_matches %>% 
  mutate(w_hgt_avg = (w_p1_hgt + w_p2_hgt)/2, 
         l_hgt_avg = (l_p1_hgt + l_p2_hgt)/2,
         tall_win = case_when(w_hgt_avg > l_hgt_avg ~ 1,
                              w_hgt_avg < l_hgt_avg ~ -1,
                              TRUE ~ 0),
        tall_win_sd = case_when((w_hgt_avg - l_hgt_avg) > hgt_sd  ~ 1,
                    (w_hgt_avg - l_hgt_avg) < -hgt_sd  ~ -1,
                    TRUE ~ 0)) %>% 
  select(circuit:gender, bracket, round, w_hgt_avg:tall_win_sd)

df %>% skim()
```

The tournament round is a character. Let's convert to a numeric.

```{r}
df <- df %>% 
  mutate(round = as.numeric(str_remove(round,"Round ")))
```


For simplicity, and becuase the amount of missing data isn't huge, we're going to just drop NAs.
```{r}
df <- df %>% drop_na()

df %>% skim()
```


```{r}
df %>% glimpse()

df_input <- df %>% 
  mutate(gender = if_else(gender =="M", "Mens", "Womens")) %>% 
  filter(tall_win != 0) %>% 
  group_by(gender, tournament) %>% 
  count(tall_win) %>% 
  mutate(freq = n / sum(n)) %>% 
  filter(tall_win == 1)

df_input %>% glimpse
```

```{r}
p_ridges_hgt <- ggplot(data = df_input, 
  aes(x = freq, y = gender, fill = ..x..)) +
  geom_density_ridges_gradient() +
  scale_fill_viridis(name = "Frequency", option = "D") +
  xlab(label = "Winning Percentage ") +
  ylab(label = "Gender") +
  ggtitle(label = "Height Advantage in Beach Volleyball",
          subtitle = "Distribution of Tournament Winning Percentage for Taller Teams") +
  theme_ridges() + 
  theme(legend.position = "none")

p_ridges_hgt
ggsave(path = "output/", filename = "ridges_plot.png", plot = p_ridges_hgt, width = 12, height = 9, units = "in")
```