---
title: '#TidyTuesday - Volcano Eruptions'
author: "Evan Canfield"
date: "5/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Link**: [https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-05-12](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-05-12)

# Load
## Libraries
```{r load_library}
if (!require(pacman)) {install.packages('pacman')} 
p_load(
  animation,
  data.table,
  gganimate,
  ggthemes,
  Hmisc,
  lubridate,
  maps, 
  mapdata,
  maptools,
  plotly,
  rgdal,
  rgeos,
  rnaturalearth,
  sf,
  skimr,
  tidyverse,
  viridis
)
```

## Data
```{r load_data}
volcano <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')
eruptions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/eruptions.csv')
events <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/events.csv')
tree_rings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/tree_rings.csv')
sulfur <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/sulfur.csv')
```

# EDA

```{r}
volcano %>% head()

volcano %>% skim()
```

```{r}
eruptions %>% head()
```

```{r}
events %>% head()
```

```{r}
tree_rings %>% head()
```

```{r}
sulfur %>% head()
```

```{r}
events %>% 
  group_by(event_type) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n))
```


# Exploratory Visuals
## Tree Rings
```{r}
ggplot(data = tree_rings, 
       mapping = aes(x= n_tree, y = europe_temp_index, color = year)) +
  geom_point() +
  theme_minimal()
```

* Determine Century
* Drop year 2000. Only year in it's century
* Map color to century. See if there are any patterns
```{r}
tree_rings %>% 
  filter(year != 2000) %>% 
  mutate(century = trunc(year/100)) %>% 
  ggplot(mapping = aes(x= n_tree, y = europe_temp_index, color = century)) +
  geom_point() +
  scale_color_viridis() +
  theme_minimal()
```

* Take averages per century
* Inspect for outliers.

```{r}
p <- tree_rings %>% 
  filter(year != 2000) %>% 
  mutate(century = trunc(year/100)) %>% 
  group_by(century) %>% 
  summarise(n_tree_avg = mean(n_tree),
            europe_temp_index_avg = mean(europe_temp_index)) %>% 
  ggplot(mapping = aes(x= n_tree_avg, y = europe_temp_index_avg, color = century)) +
  geom_point() +
  scale_color_viridis() +
  theme_minimal()

ggplotly(p)
```


# Visual
The basic visual I want to make is a map, perhaps an animation. The map should center the Ring of Fire. 

First we need to create our core dataset. 

```{r}
volcano %>% glimpse()

eruptions %>% glimpse()

df_input <- eruptions %>% 
  mutate(eruption_date = ymd(paste0(start_year,"-",start_month,"-",start_day))) %>% 
  select(volcano_number:eruption_category, vei, eruption_date, start_year, evidence_method_dating, latitude, longitude) %>% 
  mutate(year_ago = 2020 +1-start_year)

df_input
```

Look at how vocano eruptions are documented over the years.
```{r}
df_input %>% 
  ggplot(aes(x = year_ago, fill = eruption_category)) +
  geom_histogram() +
  scale_x_log10() +
  theme_minimal()
```

Import a country based world map from **rnaturalearth**.
```{r}
NE_countries <-  ne_countries(returnclass='sp')

class(NE_countries)
```


```{r}
# shift central/prime meridian towards west - positive values only
shift <- 180+30

# create "split line" to split country polygons
WGS84 <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

split.line = SpatialLines(list(Lines(list(Line(cbind(180-shift,c(-90,90)))), ID="line")), 
                          proj4string=WGS84)

# NOTE - in case of TopologyException' errors when intersecting line with country polygons,
# apply the gBuffer solution suggested at:
# http://gis.stackexchange.com/questions/163445/r-solution-for-topologyexception-input-geom-1-is-invalid-self-intersection-er
# NE_countries <- gBuffer(NE_countries, byid=TRUE, width=0)

# intersecting line with country polygons
line.gInt <- gIntersection(split.line, NE_countries)

# create a very thin polygon (buffer) out of the intersecting "split line"
bf <- gBuffer(line.gInt, byid=TRUE, width=0.000001)  

# split country polygons using intersecting thin polygon (buffer)
NE_countries.split <- gDifference(NE_countries, bf, byid=TRUE)

# plot(NE_countries.split) # check map
class(NE_countries.split) # is a SpatialPolygons object

```


```{r}
# create a bounding box - world extent
b.box <- as(raster::extent(-180, 180, -90, 90), "SpatialPolygons")

# assign CRS to box
proj4string(b.box) <- WGS84

# create graticules/grid lines from box
grid <- gridlines(b.box, 
                  easts  = seq(from=-180, to=180, by=20),
                  norths = seq(from=-90, to=90, by=10))

# create labels for graticules
grid.lbl <- labels(grid, side = 1:4)

# transform labels from SpatialPointsDataFrame to a data table that ggplot can use
grid.lbl.DT <- data.table(grid.lbl@coords, grid.lbl@data)

# prepare labels with regular expression:
# - delete unwanted labels
grid.lbl.DT[, labels := gsub(pattern="180\\*degree|90\\*degree\\*N|90\\*degree\\*S", replacement="", x=labels)]

# - replace pattern "*degree" with "°" (* needs to be escaped with \\)
grid.lbl.DT[, lbl := gsub(pattern="\\*degree", replacement="°", x=labels)]

# - delete any remaining "*"
grid.lbl.DT[, lbl := gsub(pattern="*\\*", replacement="", x=lbl)]

# adjust coordinates of labels so that they fit inside the globe
grid.lbl.DT[, long := ifelse(coords.x1 %in% c(-180,180), coords.x1*175/180, coords.x1)]
grid.lbl.DT[, lat  := ifelse(coords.x2 %in% c(-90,90), coords.x2*82/90, coords.x2)]
```


```{r}
# give the PORJ.4 string for Eckert IV projection
PROJ <- "+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" 


# transform graticules from SpatialLines to a data table that ggplot can use
grid.DT <- data.table(map_data(SpatialLinesDataFrame(sl=grid, 
                                                     data=data.frame(1:length(grid)), 
                                                     match.ID = FALSE)))
# project coordinates
# assign matrix of projected coordinates as two columns in data table
grid.DT[, c("X","Y") := data.table(project(cbind(long, lat), proj=PROJ))]

# project coordinates of labels
grid.lbl.DT[, c("X","Y") := data.table(project(cbind(long, lat), proj=PROJ))]

# transform split country polygons in a data table that ggplot can use
Country.DT <- data.table(map_data(as(NE_countries.split, "SpatialPolygonsDataFrame")))

# Shift coordinates
Country.DT[, long.new := long + shift]
Country.DT[, long.new := ifelse(long.new > 180, long.new-360, long.new)]

# project coordinates 
Country.DT[, c("X","Y") := data.table(project(cbind(long.new, lat), proj=PROJ))]
```


We need to perform the same shift and projection of coordinates to the volcano data.
```{r}
# Shift coordinates
df_input <- df_input %>% 
  mutate(long.new = longitude + shift,
         long.new = if_else(long.new > 180, long.new -360, long.new))

# project coordinates 
eruption_proj <- as.data.frame(project(cbind(df_input$long.new, df_input$latitude), proj=PROJ))

df_input <- df_input %>% mutate(long_proj  = eruption_proj$V1,
                            lat_proj = eruption_proj$V2)

df_input
```

## Blank Map - White / Gray
```{r}
p_map_white_grey <- ggplot() + 
  
  # add projected countries
  geom_polygon(data = Country.DT, 
               aes(x = X, 
                   y = Y, 
                   group = group), 
               colour = "gray90", 
               fill = "gray80", 
               size = 0.75) +
  
  # add graticules
  # geom_path(data = grid.DT, 
  #           aes(x = X, y = Y, group = group), 
  #           linetype = "dotted", colour = "grey50", size = .25) +
  
  # add a bounding box (select graticules at edges)
  geom_path(data = grid.DT[(long %in% c(-180,180) & region == "NS")
                           |(long %in% c(-180,180) & lat %in% c(-90,90) & region == "EW")], 
            aes(x = X, y = Y, group = group), 
            linetype = "solid", colour = "black", size = .3) +
  
  # # add graticule labels
  # geom_text(data = grid.lbl.DT, # latitude
  #           aes(x = X, y = Y, label = lbl), 
  #           colour = "grey50", size = 2) +
  
  # ensures that one unit on the x-axis is the same length as one unit on the y-axis
  coord_equal() + # same as coord_fixed(ratio = 1)
  
  # set empty theme
  theme_void()

p_map_white_grey
```


## Blank Map - Black / Gray
```{r}
p_map_black_grey <- ggplot() + 
  
  # add a bounding box (select graticules at edges)
  geom_polygon(data = grid.DT[(long %in% c(-180,180) & region == "NS")
                           |(long %in% c(-180,180) & lat %in% c(-90,90) & region == "EW")], 
            aes(x = X, y = Y, group = group), 
            linetype = "solid", colour = "black", size = .3) +
  
  # add projected countries
  geom_polygon(data = Country.DT, 
               aes(x = X, 
                   y = Y, 
                   group = group), 
               colour = "gray90", 
               fill = "gray90",
               size = 0.25) +
  
  # add graticules
  # geom_path(data = grid.DT, 
  #           aes(x = X, y = Y, group = group), 
  #           linetype = "dotted", colour = "grey50", size = .25) +
  
  
  # # add graticule labels
  # geom_text(data = grid.lbl.DT, # latitude
  #           aes(x = X, y = Y, label = lbl), 
  #           colour = "grey50", size = 2) +
  
  # ensures that one unit on the x-axis is the same length as one unit on the y-axis
  coord_equal() + # same as coord_fixed(ratio = 1)
  
  # set empty theme
  theme_void()

p_map_black_grey
```


# Map with Geom Point Eruptions
```{r}
cutoff_year = 1800
point_alpha = 0.25
point_size = 4
point_shape = 16
point_color = '#FF0000'

theme_erupt <- function(){
  theme(
    plot.title = element_text(size = 30),
    plot.subtitle = element_text(size = 25),
    plot.caption = element_text(size = 20),
    plot.margin =  margin(0.5, 0.5, 0.5, 0.5, "cm")
  )
}


p_static <- p_map_white_grey + 
  geom_point(data = df_input %>% filter(start_year >= cutoff_year),
               mapping = aes(x = long_proj, 
                             y = lat_proj),
             color = point_color, 
             alpha = point_alpha,
             size = point_size,
             shape = point_shape) +
  ggtitle(label = "Vocanic Eruptions Around the World",
          subtitle = paste0("From ", cutoff_year," Through Today")) +
  labs(caption = "#TidyTuesday || Created By: Evan Canfield") + 
  theme_erupt()

ggsave(filename = "output/featured_image.png", plot = p_static, width = 16, height = 9, units = "in")

p_static
```


```{r}
p_animate <- p_static +
  transition_manual(eruption_date, cumulative = TRUE) +
  ease_aes('bounce-in') +
  enter_fade() +
  exit_fade()

# Animation Time (sec)
animation_time = 60

# Frames per Second
fps = 10

height_opt = 900
width_opt = 1600
dim_factor = 2

eruption_gif <- animate(p_animate, 
        nframes = animation_time * fps,
        fps = fps,
        height = height_opt/dim_factor,
        width = width_opt/dim_factor)
```

```{r}
eruption_gif
anim_save(filename = "output/eruptions.gif", animation = eruption_gif )
```

